Nanophotonics with optical phonons in two dimensions

### 2018-11-29 paper.tex
wrote the last section on emitter strong coupling to phonons, defined scope of the rest of paper

\begin{abstract}
Extreme confinement of electromagnetic energy by phonon-polaritons promises extremely strong and novel forms of control over the dynamics of matter. To bring such control to its ultimate limit, it is important to consider phonon-polaritons in two dimensional systems. Recent studies have pointed out that in two-dimensional systems, splitting between longitudinal and transverse (LO and TO) optical phonons is absent at the $\Gamma$ point.  Does this lack of LO-TO splitting imply the lack of a strongly confined electromagnetic (phonon-polariton) mode in polar monolayers? Here, we derive a universal form for the spatially-nonlocal dielectric function of a polar monolayer specified by: the TO phonon frequency, the non-zero group velocity of the LO phonon, and the phonon lifetime. Our analysis reveals that the phonon-polariton of the bulk is ``replaced" by the LO phonon. We present the confinement and propagation losses of LO phonons, finding that high LO phonon confinement and reasonable propagation quality factors coincide in regions which are difficult to detect with near-field optical microscopy, but may be more amenable to electron energy-loss spectroscopy. Finally, we show that 2D LO phonons in hBN present extremely strong light-matter interactions with external emitters, providing possibilities for 10,000,000-fold enhanced spontaneous emission and strong coupling with vibrational emitters.
\end{abstract}

@@

In this Letter, we develop core results in the nascent field of optical phononics in two-dimensional materials by deriving a universal form for the dielectric function of a polar monolayer, incorporating spatially nonlocal effects, which is specified in fully in terms of the dispersion relation of LO and TO phonons in two dimensions. Using these parameters for important monolayers such as hexagonal boron nitride, we present the confinement and propagation losses of these modes, identifying the frequency region where they should be most easily detected. We then show that these modes essentially provide the ultimate limit of light-matter interaction between emitters and polar materials, showing that for atom-like emitters, their spontaneous decay can be enhanced seven orders of magnitude by the emitter-LO phonon coupling. For free-electrons, we show a substantial probability of multiple LO phonon excitation can occur.

@@

In Fig. 3a, we show the enhancement (or Purcell factor) $F_p$ of spontaneous emission of 2D LO phonons at frequency $\omega_0$ by an external emitter at different distances $d$ away from the surface. The rate of spontaneous emission $\Gamma$ into 2D LO phonons obeys
\begin{equation}
\Gamma = \frac{3c^3}{2\omega_0^3}\int dq~q^2e^{-2qz}\text{Im }r_p(q,\omega_0) \times \Gamma_0 \equiv F_p(\omega_0)\Gamma_0,
\end{equation}
where $\Gamma_0$ is the free-space spontaneous emission rate. As can be seen from Fig. 3a, for an emitter position 5 nm away from the surface, the Purcell factor can be very high, corresponding to over a million-fold increase in the rate of 2D LO phonon emission compared to photon emission into free-space. While this rate-enhancement is already quite high, it is a natural question what the highest possible spontaneous-emission rate enhancement is as a function of frequency. Recently, material-geometry-independent limits for spontaneous emission enhancement have been derived. In particular, the maximum rate of spontaneous emission enhancement is bounded by
\begin{equation}
...
\end{equation}
This is plotted in Fig. 3b for the case of monolayer hBN.

Finally, we address one manner of externally probing 2D LO phonons without the need for SNOM or emitters: with electron energy losses in free electrons. As the energy losses under consideration will be on the order of 200 meV for single phonon excitation in monolayer hBN, the electron energy spread which determines the width of EELS peak (including the zero-loss peak) must be substantially smaller than 200 meV. Fortunately, this is possible in high-resolution EELS, which is particularly useful for probing surface phenomena, and has been applied to measure the dispersion relation of plasmons in 2D or quasi-2D systems. Through advances in monochromation, electrons have also been recently used to probe optical phonons in bulk. In Fig. 4, we show the EELS probability for electrons with energies on the order of 100 keV traversing a polar monolayer . Because of the strong coupling between matter and LO phonons shown in Fig. 3, one expects multiple LO phonon excitation to be possible. Thus, this effect must be taken into account in the EELS. 
\begin{equation}
...
\end{equation}

>>>>

\begin{abstract}
Extreme confinement of electromagnetic energy by phonon-polaritons promises extremely strong and novel forms of control over the dynamics of matter. To bring such control to its ultimate limit, it is important to consider phonon-polaritons in two dimensional systems. Recent studies have pointed out that in two-dimensional systems, splitting between longitudinal and transverse (LO and TO) optical phonons is absent at the $\Gamma$ point.  Does this lack of LO-TO splitting imply the lack of a strongly confined electromagnetic (phonon-polariton) mode in polar monolayers? Here, we derive a universal form for the spatially-nonlocal dielectric function of a polar monolayer specified by: the TO phonon frequency, the non-zero group velocity of the LO phonon, and the phonon lifetime. Our analysis reveals that the phonon-polariton of the bulk is ``replaced" by the LO phonon. We present the confinement and propagation losses of LO phonons, finding that high LO phonon confinement and reasonable propagation quality factors coincide in regions which are difficult to detect with current near-field optical microscope techniques. We conclude by studying the interaction of external emitters with two-dimensional hBN nanostructures, finding extreme enhancement of spontaneous emission due to 2D phonon channels, and the possibility of multi-mode strong and ultra-strong coupling between an external emitter and hBN phonons.  \end{abstract}

@@

In this Letter, we develop core results in the nascent field of optical phononics in two-dimensional materials by deriving a universal form for the dielectric function of a polar monolayer, incorporating spatially nonlocal effects, which is specified in fully in terms of the dispersion relation of LO and TO phonons in two dimensions. Using these parameters for important monolayers such as hexagonal boron nitride, we present the confinement and propagation losses of these modes, identifying the frequency region where they should be most easily detected. We then show that these modes essentially provide the ultimate limit of light-matter interaction between emitters and polar materials, showing that for atom-like emitters, their spontaneous decay can be enhanced seven orders of magnitude by the emitter-LO phonon coupling, which due to the narrow bandwidth of phonon-polariton resonances, can lead to the possibility of realizing multi-mode strong coupling between an external emitter and optical phonons in hexagonal boron nitride.

@@

In Fig. 3a, we show the enhancement (or Purcell factor) $F_p$ of spontaneous emission of 2D LO phonons at frequency $\omega_0$ by an external emitter at different distances $d$ away from an infinite sheet of two-dimensional boron nitride. The rate of spontaneous emission $\Gamma$ into 2D LO phonons obeys
\begin{equation}
\Gamma = \frac{3c^3}{2\omega_0^3}\int dq~q^2e^{-2qz}\text{Im }r_p(q,\omega_0) \times \Gamma_0 \equiv F_p(\omega_0)\Gamma_0,
\end{equation}
where $\Gamma_0$ is the free-space spontaneous emission rate. As can be seen from Fig. 3a, for an emitter position 5 nm away from the surface, the Purcell factor can be very high, corresponding to over a million-fold increase in the rate of 2D LO phonon emission compared to photon emission into free-space. While this rate-enhancement is already quite high, it is a natural question how this coupling can be improved by nanostructuring, which will take the local density of electromagnetic states of Fig. 3a and concentrate it around frequencies associated with resonances of the nanostructure.For simplicity, we consider the effect of nanostructuring the hBN into a disk, which leads to the formation of modes as in Fig. 3b, in which we plot the surface charge density associated with each resonant mode. In Fig. 3c, we plot the enhancement of the spontaneous emission rate for an external emitter polarized perpendicularly to the plane of the disk and placed 5 nm away from the surface of the disk. We find that the rate of spontaneous emission of 2D optical phonons is approximately 8 orders of magnitude larger than the rate of spontaneous emission of photons in the far field. For an infrared emitter at a transition wavelength of 7 $\mu$m with a free-space radiative lifetime of 1 $\mu$s, 5 nm away from an hBN disk, the coupling rate to 2D optical phonons would be on the same scale as the optical phonon frequency itself (about $2 \times 10^{14}$ $ s^{-1}$). This coupling rate thus implies coupling between an emitter and the field in the regime of ultra-strong coupling. Thus, the extreme confinement of electromagnetic energy associated with LO phonons in two dimensions enables the possibility of realizing ultrastrong coupling of an atom or molecule with optical phonons in a polar material, allowing the potential realization of new coupled states of quantum emitters and phonons.
